---
output: word_document
---

# Preprocessing

All samples, whether from GEO or a batch of diagnostic samples from our own lab, have to be *preprocessed* as the first step.
Shown below is the code for preprocessing our own diagnostic samples. 
The only difference between the code below and how we process GEO samples is that we used the 
IDAT files from GEO series GSE90496 to derivee batch effect coefficients for each material, FFPE or Frozen.
For our own samples, the batch effects are then applied to adjust for material specific noise. 
The end result of this step is a Mathyl Set and corresponding Beta values for each sample in the batch. 

```{r cache=TRUE, eval=FALSE}

preprocessDiagnosticSamples <- function (diag_id, material = "FFPE") {

    # get basenames for idat files
    data_dir = here::here("data", diag_id)
    basenames <- unique(here::here(data_dir, 
        gsub("_Grn.*", "", 
            gsub("_Red.*", "", 
                list.files(path = data_dir, pattern = "*.idat")
            )
        )
    ))

    # read idat files into an RGSet
    rgset <- read.metharray(basenames, verbose=TRUE, force=TRUE)

    # preprocess RGSet and convert to a MethylSet
    mset <- MNPpreprocessIllumina(rgset)
    rownames(mset) <- sub("_.*", "", rownames(mset))

    # apply probe filtering to the MethylSet and save
    mset <- getFilteredMethylSet(mset)
    saveRDS(mset, file=here::here("results", paste0(diag_id,"_mset.rds")))

    methy <- getMeth(mset)
    unmethy <- getUnmeth(mset)

    # apply batch effects from saved coefs
    coefs <- loadSavedCoefs(REF_GSE_ID)
    ba <- applyBatchEffectCoefs(methy, unmethy, material, coefs)

    # recalculate betas using the Illumina Genome Studio offset and save
    illumina_offset <- 100
    betas <- ba$methy.ba / (ba$methy.ba + ba$unmethy.ba + illumina_offset)
    betas <- as.data.frame(t(betas))
    saveRDS(betas, file=here::here("results", paste0(diag_id, "_betas.rds")))
}

```

From above, the getFilteredMethylSet() function is responsible for removing probes that are not common to 
all of 450K, EPIC, and EPICv2 arrays. 
These common probes are formed from the intersection of probes from the annotation object for each version. 
By doing this, we are able to classify tumor samples from EPICv2 arrays using a random forest model derived
from samples from 450K arrays. While this is flexible, it does mean that the extra probes added to the EPIC 
arrays are not used. 

```{r cache=TRUE, eval=FALSE}

library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(here)

# get the probes from the annotation objects
anno_450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno_epic <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
anno_epicv2 <- getAnnotation(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
probes_450k <- rownames(anno_450k)
probes_epic <- rownames(anno_epic)
probes_epicv2 <- rownames(anno_epicv2)

probes_epicv2 <- sub("_.*", "", probes_epicv2)

# intersect all probes
valid_probes1 <- intersect(probes_450k, probes_epic)
valid_probes2 <- intersect(valid_probes1, probes_epicv2)

writeLines(valid_probes2, con=here("probes", "common_450k_epic_epicv2.txt"))
```

# Training 

This section only applies to the GEO series GSE90496 samples, since they are used to derive the random forest for model.
The main line code for this is shown below. We first load the Betas for GSE90496, filter the top 100K most variably
methylated probes, run the random forest algorithm in *Importance* mode, and finally filter the top 30K
most important probes. These filtered probes are then used to train the random forest model used for classification. 
The end result of training is a saved random forest model for classication of CNS tumors. 

```{r cache=TRUE, eval=FALSE}

library(here)

source(here("R","loading.R"))
source(here("R","training.R"))
    
betas <- loadSavedBetas("GSE90496")
anno <- loadSavedAnno("GSE90496")

y <- as.factor(anno$`methylation class:ch1`)

betas <- getFilteredBetasBasedOnImportance(betas, y)

rf_model <- getRandomForestModel(betas, y)

saveRDS(rf_model, file=file.path("results", paste0(REF_GSE_ID,"_model.rds")))
```

To implement the random forest algorithm, we use the *ranger* package. Below, is where we filter out
probes for training the random forest model based on variance and importance.  

```{r cache=TRUE, eval=FALSE}

getFilteredBetasBasedOnImportance <- function (betas, y) {

    N_CORES <- 8
    N_TREES <- 10000
    N_TOP_BETAS <- 100000 # top variably methylated probes
    N_TOP_FEATURES <- 30000 # top feature importance
    
    seed <- 180314
    set.seed(seed,kind ="L'Ecuyer-CMRG")

    # first filter for the most varaibly methylated probes
    betas <- betas[,order(-apply(betas,2,sd))[1:N_TOP_BETAS]]
    
    rf_varsel <- ranger(
            x=betas,
            y=y,
            num.trees=N_TREES,
            num.threads=N_CORES,
            importance="permutation")

    # get permutation variable importance
    or <- order(rf_varsel$variable.importance, decreasing=TRUE)

    # finally, filter for the most import variables determined by ranger 
    betas <- betas[,or[1:N_TOP_FEATURES]]

    return(betas)
}

```

Finally, here is the code for training the final random forest model. 

```{r cache=TRUE, eval=FALSE}

getRandomForestModel <- function (betas, y) {
    
    N_CORES <- 8
    N_TREES <- 10000
    
    seed <- 180314
    set.seed(seed,kind ="L'Ecuyer-CMRG")
    
    rf_model <- ranger(
            x=betas,
            y=y,
            probability=TRUE,
            num.trees=N_TREES,
            num.threads=N_CORES,
            keep.inbag=TRUE,
            verbose=TRUE)

    return(rf_model)
}

```

# Classification

Once we have the trained random forest model from above along with preprocessed diagnostic samples, it's time to perform the classification of CNS tumors based on methlaytion profiles. 
Below, given an diagnostic samples batch id or a GEO series id, samples classes are predicted. 
Once the raw scores are obtained from the model, we perform score callibration using a multinomial logistic regression. 

```{r cache=TRUE, eval=FALSE}

predictSampleScores <- function(id) {

	# load betas for sample
	betas <- loadSavedBetas(id)

	# load rf model from reference data
	rf_model <- loadSavedModel(REF_GSE_ID)

	# predict raw scores from rf model
	rf_vars <- rf_model$forest$independent.variable.names
	filtered_betas <- betas[match(rf_vars, colnames(betas))]
	p <- predict(rf_model, data = filtered_betas)
	scores <- p$predictions
	rownames(scores) <- rownames(filtered_betas)

	# calibrate scores using calfit model
	calfit <- loadSavedCalfit(REF_GSE_ID)
	scores <- predict(
			calfit$glmnet.fit,
			newx = scores,
			type = "response",
			s = calfit$lambda.1se
		)[,, 1] # use lambda estimated by 10fold CVlambda

	return(scores)
}

```

Once the scores are obtained as above, they have to be reshaped using Tidy data techniques for the final Excel output. 

```{r cache=TRUE, eval=FALSE}

evaluateDiagnosticSampleScores <- function (scores) {

	scores_df <- as.data.frame(scores, row.names = rownames(scores))
	scores_t <- tibble::rownames_to_column(scores_df, var = "sample_id")
	
	scores_mc <- scores_t |>
		pivot_longer(
			cols = !(sample_id),
			names_to = "mc",
			values_to = "mc_score"
		) |>
		group_by(sample_id) |>
		slice_max(mc_score) 

	scores_mcf <- scores_t |>
		pivot_longer(
			cols = !(sample_id),
			names_to = "mc",
			values_to = "mc_score"
		) |>
		mutate(
			mcf = mcf_lookup(mc),
			.after = mc
		) |>
		group_by(sample_id, mcf) |>
		summarize(
			mcf_score = sum(mc_score),
			.groups = "drop_last"
		) |>
		slice_max(order_by = mcf_score)

	scores_j <- scores_mc |> 
		inner_join(scores_mcf, join_by(sample_id)) |>
		mutate(
			mc_descr = MC[mc]
		)

	return (scores_j)
}

```


